Supplementary Information

### 2020-08-27 doc/manuscript/MEE_manuscript.Rmd
first pass at detailing the methods
I tried to improve the description of the GLS models and be more specific about which responses variables are analysed in the different steps.
There is still some work but I think we cam go ahead with first round of coauthor reviews.
Doing a pull request so that you can check what I changed before accepting. and in case you've made changes while I was editing. This is the the .Rmd document, I didn't knit it.

1. Tree rings provide a valuable long-term record for understanding how climate shapes forest productivity. However, traditional analysis methods have not been designed to simultaneously account for the effects of tree size and climate, which has limited the potential to use tree-rings to understand forest productivity in the current era of rapid climate change.

2. Here, we develop a new method that allows simultaneous non-linear modeling of the effects of objectively determined principle climate drivers and tree diameter. 
Specifically, we first identify the most important climate drivers using the climwin R package.
We then include these in generalized least squares models that simultaneously fit the detrending splines needed to pull out climate signals and flexibly account for nonlinearity of responses to tree diameter and climate variables.
We apply this method to tree-ring data from **#** species at ten globally distributed sites spanning a wide range of forest types.

@@

Once $DBH$ had been reconstructed, we calculated aboveground biomass growth increments ($\Delta AGB$) based on $DBH$ and $\Delta r$.
Specifically, for each year $y$, we used biomass allometries to estimate $ABG_y$ based on $DBH_y$ and $AGB_{y+1}$ based on [$DBH_y + \Delta r$]. 
We then calculated $\Delta AGB_y$ as [$AGB_{y+1}-AGB_y$]. 
For temperate sites, biomass allometries were sourced from allo-db (DETAILS; Gonzalez-Akre et al. in prep).
For tropical sites, biomass allometries were sourced from the BIOMASS package (REF).

@@

Our analysis consisted of two main steps: (1) identification of the most important climate drivers, and (2) combining $DBH$ and climate drivers into a multivariate model (Fig. **1**). The analysis was run separately for each site.

@@

First, we identified the most important climate driver for each of two categories of variables, all at the monthly timescale: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation). 
Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
Specifically, we used a generalized additive model (GAM) to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each variable group. 
*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
We began by verifying that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix S2. (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))

We then used *climwin* to identify the best climate variables across the full set of cores from each site, noting that whereas our goal was to identify the most significant climate drivers at a site level, identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*). (**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51))
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models.

@@

Second, we combined candidate climate drivers and reconstructed $DBH$ data in a GLS model (Fig. **1**). 
*(need to describe GLS models better)*
*(DBH not included in models focused on climate sensitivity.)*
We checked for collinearity among the full set of variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3. 
For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), a spline of $DBH$ (3 knots), a random intercept for each tree, and *a temporal autocorrelation matrix*. 
We considered the model with lowest AIC as the top model. 
We then tested the individual significance of each term in the top model, including both first- and second-order terms of polynomial fits to climate variables. To do so, we compared the AIC of the top model (containing the term) to a model lacking the term, counting as significant any term with $\Delta AIC \le$ 2.0.

>>>>

1. Tree rings provide a valuable long-term record for understanding how climate shapes forest productivity. However, traditional analysis methods have not been designed to simultaneously account for the effects of tree size and climate, because they aggregate tree ring records of multiple trees into residual chronologies. This has limited the potential to use tree-rings to understand forest productivity in the current era of rapid climate change.

2. Here, we develop a new method that allows to keep the raw tree ring measurements and individual tree's metrics to simultaneously model non-linear effects of objectively determined principle climate drivers and e.g. tree diameter. 
Specifically, we first identify the most important climate drivers and their appropriate time window of influence using the climwin R package.
We then include these in generalized least squares models to model tree rings (or derived measurements) while accounting for the temporal autocorrelation inherent to each individual tree's growth.
We apply this method to tree-ring data from **#** species at ten globally distributed sites spanning a wide range of forest types.

@@

Once $DBH$ had been reconstructed, we used biomass allometries to estimate the corresponding aboveground biomass and diameter to area equation to get the corresponding basal area. We then calculated aboveground biomass growth increments ($\Delta AGB$) as [$AGB_{y+1}-AGB_y$] and basal area increment ($BAI$) as [$BAI_{y+1}-BAI_y$].
For temperate sites, biomass allometries were sourced from allo-db (DETAILS; Gonzalez-Akre et al. in prep).
For tropical sites, biomass allometries were sourced from the BIOMASS package (REF).

@@

Our analysis consisted of two main steps: (1) identification of the most important climate drivers, and (2) combining $DBH$ and climate drivers into a multivariate model (Fig. **1**). The analysis was run separately for each site and each response variable ($\Delta r$, $BAI$, or $\Delta AGB$).

@@

First, we identified the most important climate driver for each of two categories of variables, all at the monthly timescale: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation). 
For this, we detrended the response variable to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS). Specifically, we used a generalized additive model (GAM) to fit a spline to individual tree growth records ($\Delta r$, $BAI$, or $\Delta AGB$), thereby producing residuals.
We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each variable group. 
*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $\Delta r$, $BAI$, or $\Delta AGB$.
We specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*). (**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51)**)
*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
We veryfied  that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix S2. (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)**)
We note that whereas our goal was to identify the most significant climate drivers at a site level, identification of the top climate drivers for individual species would be optimized by analyzing each species separately.

@@

Second, we combined candidate climate drivers identified in the previous step in a GLS model (Fig. **1**). 
For this step, our response variables are not the residuals of detrended $\Delta r$, $BAI$, or $\Delta AGB$ but instead $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$ because  $\Delta r$, $BAI$, or $\Delta AGB$ are log-linear. The temporal autocorrelation of individual tree's growth is accounted for by the specifying an autocorrelation structure of order 1, with $Year$ as a continuous time covariate and $coreID$ as a grouping factor, in the GLS's model specification.
For each species (at each site) independently, we ran every combination of the candidate climate variables and $DBH$ (for the subset of cores with successfully reconstructed $DBH$), including both first- and second-order terms of polynomial for each.
We considered the model with lowest AIC as the top model. 
The top model was used to visualize the effects of each variables while the others are kept constant at the median in the various figures of this manuscript. For these visualizing tools, we also fetched the t-test's p-value of each term of the model and represented the significance using different line types and the model predictions represented were back-transformed to show the effect on the orginal sclae of $\Delta r$, $BAI$, or $\Delta AGB$.
Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3 (which was none).

*(DBH not included in models focused on climate sensitivity.)*
